Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies

### 2018-10-16 man/ms.Rmd
add comments of review

---
title: "Relict populations of a deciduous oak show high resilience to drought despite land-use legacies"
author: "AJ Perez-Luque; G. Gea-Izquierdo; ..."
output:
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    highlight: null
    reference_docx: ../../templates/template_ms.docx
bibliography: ../../refs/references.bib
csl: ../../refs/ecology.csl
---

@@

# Abstract 
* Trees showed resilience to drought events, despite the ‘a priori’ vulnerability that they can express as a result of the impact of climate change on relict stands resulting from intensive land-use in history.

* Oak resilience to drought followed a dryness gradient: Northern populations, which thrive under a drier climate, were less resilient to drought both in primary and secondary growth than Southern populations where lower altitudes were also less resilient than those at high-altitudes.

@@

Drought is a natural phenomenon occurring when water availability is significantly below-normal levels [@Dai2011]. It is one of the most frequent and widespread climatic extremes affecting ecological systems across every climatic zone worldwide [@Mishra2010; @Dai2011; @Ummenhofer2017]. Summer drought is a characteristic feature of the Mediterranean climate [@Lionello2012], but an increase in the frequency and severity of drought events have been recorded for the Mediterranean region [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported for the last three decades [@Spinoni2017b]. Climate change projections suggest that extreme events, like droughts, will become more frequent and severe [@Trenberth2014; @IPCC2013], specially in the Mediterranean area [@Spinoni2017a].

Drought causes significant changes in ecosystem productivity and water dynamics that can severely affect forests through changes in plant physiology, phenology and carbon allocation [@Allen2010; @Ummenhofer2017]. In the last decades the assessment of the drought impacts on ecosystems has gained attention [@Allen2010; @Clark2016], especially considering some studies that suggest an increase in severity, geographic extent and frequency of droughts as a result of climate change [@Dai2011; @Trenberth2014]. The response of vegetation to drought has been analyzed at different scales, from global [*e.g*. @Allen2010; @VicenteSerrano2013; @Norman2016] to local scales [see @MartinezVilalta2016 for a revision].

Vegetation responses to drought are influenced by other drivers of environmental change [@Fischer2006; @Oliver2014; @Franklin2016; @Penuelas2017], and some authors states that the impacts of drought on ecosystem must be evaluated in a context of global change considering the interactions with other drivers of environmental change [@Archaux2006; @Clavero2011], such us land-use change [@DoblasMiranda2017]. This is specially relevant for areas with a long history of landscape modification as the Mediterranean region where land-use change is a key driver of the global change [@Navarro2013; @Ameztegui2016]. For instance, @DoblasMiranda2017 in a recent review, found that land use changes and more especially drought, are crucial factors in the interactions among the drivers of global change for Mediterranean forests.

The impacts of drought are also especially relevant for populations located in the rear edge of their distribution, where climatic conditions often corresponds to the species-limits and the populations are likey to be more sensitive to small variations on climate [@Hampe2005]. The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of the species to predicted climatic changes [@Jump2010]. It becomes even more important for relict populations driven by historical land-uses, which can harbour high levels of intraspecific genetic diversity [@Valbuena2010].

Several studies analyzed the drought effects on Mediterranean tree species representing the southermost limit of their distribution [@Pasho2011; @Camarero2011; @SanchezSalguero2012; @Linares2014; @Dorado2017b; @SanchezSalguero2017; @Caminero2018], some focused on the resilience to drought of rear-edge populations [*e.g.* @SanchezSalguero2013; @Herrero2014]. Resilience is the capacity of an ecosystem, community or individual to recover pre-disturbance structure and function after a disturbance [@Holling1973; @Lloret2011; @Hodgson2015]. The assessment of forest resilience to climatic disturbances, such as extreme droughts, provides critical information about the capacity of forests to maintain their structure and to continue providing valuable ecosystem services. This is particularly relevant for populations located in their rear-edge of their distribution, especially when they are located in Mediterranean mountains.

Here we assessed the resilience to drought of rear-edge populations of a tree Mediterranean species combining several approaches: remote sensing and tree-ring [@Jump2010].

Remote sensing can be used for studying droughts impacts on ecosystems [@Deshayes2006; @Zhang2013; @AghaKouchak2015; @McDowell2015; @Norman2016]. Vegetation indices obtained from satellite, such us EVI (*Enhanced Vegetation Index*) are used as a proxy for assessing vegetation functioning and aboveground net primary production [@Huete2002], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013; @Vicca2016]. Drought impacts on vegetation using remote sensing have been analyzed at different scales [@VicenteSerrano2013; @Assal2016] and for Mediterranean area there were studies covering all region [@Gouveia2017] and others focused on local scales [@Lloret2007; @VicenteSerrano2007; @Zribi2016].

Tree-ring width is a widely used proxy for tree vitality [@Fritts1976; @Dobbertin2005; @Bhuyan2017] and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Gazol2018]. Several studies combined the use of satellite information with dendroecological data to assess the effect of droughts on vegetation [@Dorman2013; @VicenteSerrano2013; @Sanguesa2014; @Coulthard2017]; and recently, @Gazol2018 evaluated the resilience of several forests in Spain combining information derived from remote sensing and tree-ring data.

In this work we used remote sensing information and dendroecological methods to evaluate the drought impacts in both canopy greenness (as a proxy to primary growth) and tree-radial growth of a Mediterranean tree species (*Quercus pyrenaica* Willd.) located in Sierra Nevada, considered a rear-edge of their distribution. We are also interested in assessing resilience of these forest to several extreme drought episodes. We hypothesized that, relict populations driven by historical land-use at their climatic (either altitudinal or latitudinal) rear-edge are particularly vulnerable to climate change, and hence they will show low-values of resilience after several extreme droughts. Specifically, the aims of this work were: (*i*) To quantify how two recent extreme drought events influenced primary and secondary growth of *Q. pyreancia* forests in their rear edge; (*ii*) to analyze the resilience of these forests to successive extreme drought events, both in recent times and in the long-term using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions within the rear edge of the distribution of this species. In addition, within the region of study we were interested to assess whether the effect of aspect and microclimatic conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.

@@

Pyrenean oak (*Quercus pyrenaica* Willd.) forests extend through south-western France and the Iberian Peninsula reaching its southern limit in northern Morocco [@Franco1990]. In the Iberian Peninsula these forests live on siliceous soils, or soils poor in basic ions [@Vilches2014]
under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate [@RivasMartinez2002]. *Q. pyrenaica* requires over 650 mm of annual precipitation and a summer minimal precipitation between 100 and 200 mm [@MartinezParras1982; @GarciaJimenez2009], with summer rainfall being a key factor in the distribution of the species [@Gavilan2007; @delRio2007].

This species reaches its southernmost European limit in Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations of between 860 m and 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August). There are eight oak patches (2400 Has) identified (Figure 1) in this mountain range, ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys. Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010] and these populations are considered as a rear edge of the habitat distribution, which is important in determining habitat responses to expected climate change [@Hampe2005].

The populations of Pyrenean oak forests at Sierra Nevada are considered relict forests [@Melendo2000; @Vivero2000], undergoing intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2010; @Valbuena2017]. In fact, conservation status of this species for southern Spain is "Vulnerable" [@Vivero2000]. The relict presence of this species in Sierra Nevada is related both to its genetic resilience and high intraspecific genetic diversity [@Valbuena2010]. However, they are also expected to suffer the impact of climate change, due to their greater hydric requirements compared to other more drought-tolerant Mediterranean evergreen oak and pine species. Thus, simulations of the climate change effects on this habitat forecast a reduction in suitable habitats for Sierra Nevada [@Benito2011].

@@

During the second half of the XX century, several extreme drought episodes were recorded for the Iberian Peninsula (*e.g*. 1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014; @Guerreiro2017].The 2005 and 2012 drought events have been documented as two of the worst drought years in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Gouveia2014; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017] and they were characterized as extreme drought in our climatic data (Appendices S1-S3). We focused on these two drought events because limitations about temporal availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events. The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by Spinioni (see Appendix S3)

@@

Vegetation indices derived from remote sensing information have been widely used to derive descriptors of ecosystem functioning [*e.g* @AlcarazSegura2009] and they are linked with primary productivity [@Ruimy1994]. Here, vegetation greenness of *Q. pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the *Moderate Resolution Imaging Spectroradiometer* (MODIS) sensor [@Didan2015]. MODIS EVI Data (Collection 6) were obtanied using Google Earth Engine platform for the period 2000 - 2016. EVI data consits of 16-day maximun value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m. We selected the pixels covering the distribution of *Q. pyrenaica* forests in Sierra Nevada (*n* = 928 pixels). Then, a data filtering was applied to select EVI valid values. For this purpose, complementary information accompanying the EVI data was used (VI Usefulness Indices and quality flags). Values affected by high content of aerosols, clouds, snow and shadows, were filtered out following recommendations for mountain regions [@Reyes2015].

@@

To explore the effect of drought on greenness, the EVI standardized anomaly ($EVI~sa$) was computed pixel-by-pixel, since it minimizes biases in the evaluation of anomalies, providing more information about the magnitude of the anomalies [@Samanta2012; @Gao2016]. For each pixel, an annual EVI value was obtained by averaging EVI valid values. Then, the standardized anomaly was computed as:

@@

As a surrogate of mean annual primary production, annual mean EVI ($EVI_{mean}$) value was computed for each pixel for the period 2000 - 2016. $EVI_{mean}$ was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual ($r$ = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88). For exploratory purposes, temporal trends of $EVI_{mean}$ were explored at a pixel scale. The nonparametric Mann–Kendall test (Kendall, 1975; Mann, 1945) was applied using the `trend` package in R [@Thorsten2017]. The alpha level for the test was set to 0.05.

@@

### Tree sampling 
Tree sampling was carried during autumn of 2016. Trees were sampled at two locations in contrasting slopes of Sierra Nevada: San Juan (SJ), xeric site located at northern aspect; and Cáñar (CA), humid but warm site located at southern aspect (Figure 1; Table 1). For the southern site two elevations were sampled: CA-Low and CA-High. All the sites were oak monospecific and representatives of the population clusters identified for the species in this mountain range [@PerezLuque2015]. In each site between 15 and 20 dominant trees were randomly selected. Two cores of 5 mm of diameter were taken from each tree at breast heigth (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured using a girth tape and a Vertex IV ultrasonic hypsometer (Haglöf, Sweden) respectively. Stand competition of target trees was assessed by recording distance, azimuth, DBH, species and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot of 10 m radius. Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$. Then, non-parametric Kruskal-Wallis rank sum tests [@Sokal1995] were used to analyse differences between sites for height, DBH and competition indices. When significant differences were observed, we run multiple comparisons using the Dunn's-test [@Dunn1964] with Bonferroni adjustment to correct significance values. These analysis were performed uisng the `PMCMR` package in R [@Pohlert2014].

>>>>

---
title: "Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies"
author: "AJ Perez-Luque; G. Gea-Izquierdo; ..."
output:
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    highlight: null
    reference_docx: ../../templates/template_ms.docx
bibliography: ../../refs/references.bib
csl: ../../refs/ecology.csl
---

@@

# Abstract 
* Global change challenges forest ecosystems particularly in species-specific dry locations where land-use legacies have modified species composition and the forest structure as a result of intensive human pressure in the past. We analyzed forest resilience of adult trees to drought and climate change of relict Mediterranean *Quercus pyrenaica* Willd. populations in the Southern Iberian Peninsula. Forest resilience was analyzed using both dendroecological estimates of growth and MODIS satellite images as proxies of secondary and primary growth respectively.

* The studied relict populations showed expressions of management, likely reflecting the origin of the current forest structure. Both resilience indices of primary and secondary growth expressed that trees showed high resilience both to recent drought events and also to long-term climatic changes (i.e. warming), despite the ‘a priori’ vulnerability that we were expecting them to express in respone climate change on relict stands resulting from intensive land-use in history at their current rear-edge. Furthermore, the coldest stands show a positive growth trend since the end of the 1970s matching the simultaneous increase in temperature observed.

* Resilience indices showed ecological coherence along a gradient of drought severity in time. In addition, oak resilience to drought followed a dryness gradient among the three populations studied: Northern populations thriving under a drier climate were less resilient to drought both in primary and secondary growth than Southern populations where lower altitudes were also less resilient than those at high-altitudes.

* Despite being considered relict stands at the dry rear-edge for the species, they showed high resilience in response to drought. Yet, the high sensitivity to water availability expressed and positive growth trends only in the coldest and more humid stands suggest divergent response to a changing climate along very short climatic gradients and suggest that dynamics (i.e. not only adult performance but also regeneration) of those stands located at the lowest elevations, hence the altitudinal rear-edge, need to be monitored under future expected drier and warmer climatic scenarios and, particularly, if high management pressure remains.

@@

Drought is a natural phenomenon occurring when water availability is significantly below-normal levels [@Dai2011]. It is one of the most frequent and widespread climatic extremes affecting ecological systems across every climatic zone worldwide [@Mishra2010; @Dai2011; @Ummenhofer2017]. Summer drought is a characteristic feature of the Mediterranean climate [@Lionello2012], but an increase in the frequency and severity of drought events have been recorded for the Mediterranean region [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported for at least the last three decades [@Spinoni2017b]. Climate change projections suggest that extreme events, like droughts, will become more frequent and severe [@Trenberth2014; @IPCC2013], specially in the Mediterranean area [@Spinoni2017a].

Drought causes significant changes in ecosystem productivity and water dynamics that can severely affect forests through changes in plant physiology, phenology and carbon allocation [@Allen2010; @Ummenhofer2017]. In the last decades the assessment of the drought impacts on ecosystems has gained attention [@Allen2010; @Clark2016], especially considering some studies that suggest an increase in severity, geographic extent and frequency of droughts as a result of climate change [@Dai2011; @Trenberth2014]. The functional response of vegetation to drought has been analyzed at different scales, from global [*e.g*. @Allen2010; @VicenteSerrano2013; @Norman2016] to local scales [see @MartinezVilalta2016 for a revision].

Vegetation responses to drought are influenced by other drivers of environmental change [@Fischer2006; @Oliver2014; @Franklin2016; @Penuelas2017], and some authors states that the impacts of drought on ecosystem must be evaluated in a context of global change considering the interactions with other drivers of environmental change [@Archaux2006; @Clavero2011], such us land-use change [@DoblasMiranda2017]. This is especially relevant for areas with a long history of landscape modification such as the Mediterranean region where land-use change is a key driver of the global change [@Navarro2013; @Ameztegui2016]. `REVIEW` For instance, @DoblasMiranda2017 in a recent review, found that land use changes and more especially drought, are crucial factors in the interactions among the drivers of global change for Mediterranean forests.

The impacts of drought are also especially relevant for populations located in the rear edge (either latitudinal or altitudinal along climatic gradients) of their distribution, where climatic conditions often corresponds to the species-limits and the populations are likey to be more sensitive to small variations on climate [@Hampe2005]. The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of the species to predicted climatic changes [@Jump2010]. It becomes even more important for relict populations driven by historical land-uses, which can harbour high levels of intraspecific genetic diversity [@Valbuena2010].

Several studies analyzed the drought effects on Mediterranean tree species representing the southermost limit of their distribution `REVIEW`[@Pasho2011; @Camarero2011; @SanchezSalguero2012; @Linares2014; @Dorado2017b; @SanchezSalguero2017; @Caminero2018], some focused on the resilience to drought of rear-edge populations [*e.g.* @SanchezSalguero2013; @Herrero2014]. Resilience is the capacity of an ecosystem, community or individual to recover pre-disturbance structure and function after a disturbance [@Holling1973; @Lloret2011; @Hodgson2015]. The assessment of forest resilience to climatic disturbances, such as extreme droughts, provides critical information about the capacity of forests to maintain their structure and to continue providing valuable ecosystem services. This is particularly relevant for populations located in the rear-edge of their current distribution, especially when they are located in Mediterranean mountains.

Here we assessed the resilience to drought of rear-edge populations of a tree Mediterranean species combining several approaches: remote sensing and dendroecologial estimates of growth. Remote sensing can be used for studying droughts impacts on ecosystems [@Deshayes2006; @Zhang2013; @AghaKouchak2015; @McDowell2015; @Norman2016]. Vegetation indices obtained from satellite, such us EVI (*Enhanced Vegetation Index*) are used as a proxy for assessing vegetation functioning and aboveground net primary production [@Huete2002], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013; @Vicca2016]. Drought impacts on vegetation using remote sensing have been analyzed at different scales [@VicenteSerrano2013; @Assal2016] and for Mediterranean area there were studies covering all region [@Gouveia2017] and others focused on local scales [@Lloret2007; @VicenteSerrano2007; @Zribi2016].

Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality [@Fritts1976; @Dobbertin2005; @Bhuyan2017] and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976]. Several studies combined the use of satellite information with dendroecological data to assess the effect of droughts on vegetation [@Dorman2013; @VicenteSerrano2013; @Sanguesa2014; @Coulthard2017]; and recently, @Gazol2018 evaluated the resilience of several forests in Spain combining information derived from remote sensing and tree-ring data.

In this work we used remote sensing information and dendroecological methods to evaluate the drought impacts in both canopy greenness (as a proxy to primary growth) and tree-radial growth of a Mediterranean tree species (*Quercus pyrenaica* Willd.) located in Sierra Nevada, which is considered a rear-edge of species distribution and species performance is considered to be threatened. We are also interested in assessing resilience of these forests both to several extreme drought episodes and to climate change in the long-term in the last decades. We hypothesized that, relict populations driven by historical land-use at their climatic (either altitudinal or latitudinal) rear-edge are particularly vulnerable to climate change, and hence they will show low-values of resilience after several extreme droughts. Specifically, the aims of this work were: (*i*) To quantify how two recent extreme drought events influenced primary and secondary growth of *Q. pyrenaica* forests in their rear edge; (*ii*) to analyze the long-term resilience of these forests to successive extreme drought events, both in recent times and in the long-term using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions within the rear edge of the distribution of this species to detect vulnerability to climate change. In addition, within the region of study we were interested to assess whether the effect of aspect and microclimatic conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.

@@

Pyrenean oak (*Quercus pyrenaica* Willd.) forests extend through south-western France and the Iberian Peninsula reaching its southern limit in mountains areas of northern Morocco [@Franco1990]. In the Iberian Peninsula these forests live on siliceous soils, or soils poor in basic ions [@Vilches2014] under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate [@RivasMartinez2002]. *Q. pyrenaica* requires over 650 mm of annual precipitation and a summer minimal precipitation between 100 and 200 mm [@MartinezParras1982; @GarciaJimenez2009], with summer rainfall being a key factor in the distribution of the species, which can hence be considered to have low drought tolerance compared to genuine evergreen Mediterranean taxa [@Gavilan2007; @delRio2007].

The forests of this species reaches their `REVIEW` southernmost European limit in Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations between 860 m and 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August). There are eight Pyrenean oak patches (2400 Has) identified in this mountain range (Figure 1), ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys. Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010] and these populations are considered as a rear edge of the habitat distribution, which is important in determining habitat responses to expected climate change [@Hampe2005]. For this reason, the populations of Pyrenean oak forests at Sierra Nevada are considered relict forests [@Melendo2000; @Vivero2000], having these forests undergone intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2010; @Valbuena2017]. Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" [@Vivero2000] and they are expected to suffer the impact of climate change, reducting its suitable habitats in the rear-edge due to their greater hydric requirements compared to other more drought-tolerant Mediterranean evergreen oak and pine species [@GeaIzquierdo2013].`REVIEW CITA`

@@

During the second half of the twentieth century the Iberian Peninsula suffered several extreme drought episodes (*e.g*. 1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014; @Guerreiro2017].The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Gouveia2014; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017] and they were characterized as extreme drought in our climatic data (Appendices S1-S3). We focused on these two drought events because they were those included in the period where there is availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events and analyze forest resilience to drought in a longer term. The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by @Spinoni2015 (see Appendix S3).

@@

Vegetation indices derived from remote sensing information have been widely used to derive descriptors of ecosystem functioning [*e.g* @AlcarazSegura2009] because they are linked with primary productivity [@Ruimy1994]. Here, vegetation greenness of *Q. pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the *Moderate Resolution Imaging Spectroradiometer* (MODIS) sensor [@Didan2015]. MODIS EVI Data (Collection 6) were obtanied using Google Earth Engine platform for the period 2000 - 2016. EVI data consits of 16-day maximun value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m. We selected the pixels covering the distribution of *Q. pyrenaica* forests in Sierra Nevada (*n* = 928 pixels). Then, a data filtering was applied to select EVI valid values. For this purpose, complementary information accompanying the EVI data was used (VI Usefulness Indices and quality flags). Values affected by high content of aerosols, clouds, snow and shadows, were filtered out following recommendations for mountain regions [@Reyes2015].

@@

As a surrogate of mean annual primary production, annual mean EVI ($EVI_{mean}$) value was computed for each pixel for the period 2000 - 2016. To explore the effect of drought on greenness, the EVI standardized anomaly ($EVI~sa$) was computed pixel-by-pixel, since it minimizes biases in the evaluation of anomalies, providing more information about the magnitude of the anomalies [@Samanta2012; @Gao2016]. For each pixel, an annual EVI value was obtained by averaging EVI valid values. Then, the standardized anomaly was computed as:

@@

Rather than other vegetation indices like the NDVI, $EVI_{mean}$ was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual ($r$ = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88). For exploratory purposes, temporal trends of $EVI_{mean}$ were explored at a pixel scale. The nonparametric Mann–Kendall test (Kendall, 1975; Mann, 1945) was applied using the `trend` package in R [@Thorsten2017]. The alpha level for the test was set to 0.05.

@@

### Tree sampling 
Tree sampling was carried out during autumn 2016. Trees were sampled at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at northern aspect; and Cáñar (CA), a humid but warm site located at southern aspect (Figure 1; Table 1). For the southern site two elevations were sampled: CA-Low and CA-High, which constitutes the current altitudinal tree-line in the sampled slope. All the sites were oak monospecific and representatives of the population clusters identified for the species in this mountain range [@PerezLuque2015]. In each site between 15 and 20 dominant trees were randomly selected. Two cores of 5 mm of diameter were taken from each tree at breast heigth (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured using a girth tape and a Vertex IV ultrasonic hypsometer (Haglöf, Sweden) respectively. Stand competition of target trees was assessed by recording distance, azimuth, DBH, species and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot of 10 m radius. Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$. Then, non-parametric Kruskal-Wallis rank sum tests [@Sokal1995] were used to analyse differences between sites for height, DBH and competition indices. When significant differences were observed, we run multiple comparisons using the Dunn's-test [@Dunn1964] with Bonferroni adjustment to correct significance values. These analysis were performed uisng the `PMCMR` package in R [@Pohlert2014].
